---
title: "Assignment 2: Coding Basics"
author: "Allison_Barbaro"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r echo=TRUE}
#1. #creating sequence with the format seq (from, to, by)
seq(1, 100, 4)
hundred_seq <- seq(1, 100, 4)

#2. #finding the mean and median of my sequence, which I named hundred_seq
mean(hundred_seq)
median(hundred_seq)

#3. #determining whether the mean of this sequence is greater than the median
mean(hundred_seq) > median(hundred_seq)

```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r echo=TRUE}
#5 #created vectors of student names, scores, and whether they passed or failed
student_names <- c("Ally", "Maeve", "Laura", "Aileen")
#this is a character vector
student_scores <- c(48, 75, 99, 92)
#this is a double vector
pass_fail <- c(FALSE, TRUE, TRUE, TRUE)
#this is a logical vector

df <- data.frame(student_names, student_scores, pass_fail)
gradebook <- df
gradebook

colnames(gradebook) <- c('Names', 'Scores', 'P/F')
gradebook

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: A matrix can only hold a single dataset, while data frames can hold mutliple.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r}
scores_2 <- function(x) {
 ifelse(x > 50, "TRUE", "FALSE") 
}

pass_fail <- scores_2(student_scores); pass_fail


```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer:'ifelse' worked, while 'if' and 'else' did not. 'if' and 'else' functions can only check one element in a vector at a time, while 'ifelse' can check every element in a vector. Our vector contains multiple elements. 
