---
title: 'Assignment 10: Data Scraping'
author: "Allison Barbaro"
output:
  pdf_document: default
  html_document:
    df_print: paged
geometry: margin=2.54cm
editor_options:
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r set up, echo=TRUE, message=FALSE}

#1
library(tidyverse)
library(rvest)
library(dplyr)
library(ggplot2)
library(lubridate)

getwd()

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website, echo=TRUE}

#2
the_URL <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022')

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings), with the first value being "27.6400".

```{r scrape.the.data, echo=TRUE}
#3
water.system.name <- the_URL %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text()
  
PWSID <- the_URL %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text()
  
ownership <- the_URL %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text()

max.withdrawals.mgd <- the_URL %>%
  html_nodes("th~ td+ td") %>%
  html_text()

max.withdrawals.mgd <- as.numeric(max.withdrawals.mgd)

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the average daily withdrawals across the months for 2022

```{r create.a.dataframe.from.scraped.data, echo=TRUE}

#4

Date <- c("01/2022", "05/2022", "09/2022", "02/2022", 
          "06/2022", "10/2022", "03/2022", "07/2022", 
          "11/2022", "04/2022", "08/2022", "12/2022")

df_withdrawals <- data.frame("System Name" = rep(water.system.name),
                             "PWSID" = rep(PWSID),
                             "Ownership" = rep(ownership),
                             "Month" = c("Jan", "May","Sept", 
                                         "Feb", "June", "Oct", "Mar",
                                         "July", "Nov", "Apr", "Aug",
                                         "Dec"),
                             "Year" = rep(2022),
                             "Date" = Date,
                             "Max Day Use" = max.withdrawals.mgd)

df_withdrawals$Date <- my(df_withdrawals$Date)
                            
#5
ggplot(df_withdrawals) +
  geom_line(aes(x = Date, y = Max.Day.Use)) +
  labs(x = "Date", y = "Average Daily Withdrawals", 
       title = "Average Daily Withdrawals for 2022")


```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function, echo=TRUE}

#6.
scrape.it <- function(the_PWSID, the_year){

the_website <- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=',
                                the_PWSID, '&year=', the_year))

#Scrape the data
water.system.name.1 <- the_website %>% 
  html_nodes('div+ table tr:nth-child(1) td:nth-child(2)') %>%
  html_text()
  
PWSID.1 <- the_website %>% 
  html_nodes('td tr:nth-child(1) td:nth-child(5)') %>%
  html_text()
  
ownership.1 <- the_website %>%
  html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>%
  html_text()

max.withdrawals.mgd.1 <- the_website %>%
  html_nodes('th~ td+ td') %>%
  html_text()

max.withdrawals.mgd.1 <- as.numeric(max.withdrawals.mgd.1)

#Convert to a dataframe

Month <- c('Jan', 'May', 'Sep', 'Feb', 'Jun', 
           'Oct', 'Mar', 'Jul', 'Nov', 'Apr', 'Aug', 'Dec')

df_withdrawals_function <- data.frame("System Name" = rep(water.system.name.1),
                             "PWSID" = rep(PWSID.1),
                             "Ownership" = rep(ownership.1),
                             "Month" = Month,
                             "Year" = rep(the_year),
                             "Date" = my(paste0(Month, "-", the_year)),
                             "Max Day Use" = max.withdrawals.mgd.1)
df_withdrawals_function <- arrange(df_withdrawals_function, Date)
}
```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data, echo=TRUE}
#7
dataframe.2015 <- scrape.it("03-32-010", 2015)

ggplot(dataframe.2015) +
  geom_line(aes(x = Date, y = Max.Day.Use)) +
  labs(x = "Date", y = "Average Daily Withdrawals", 
       title = "Average Daily Withdrawals for 2015") 

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data, echo=TRUE}

#8
Asheville.2015 <-scrape.it('01-11-010', 2015)
view(Asheville.2015)

Asheville.Durham <- rbind(Asheville.2015, dataframe.2015)

ggplot(Asheville.Durham) +
  geom_line(aes(x = Date, y = Max.Day.Use)) +
  labs(x = "Date", y = "Average Daily Withdrawals", 
       title = 
         "Average Daily Withdrawals for 
       Asheville and Durham in 2015") +
  facet_wrap(vars(System.Name), nrow = 2)

```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "09_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data, echo=TRUE}

#9 
the_years <- c(2010, 2011, 2012, 2013, 2014, 2015, 2016,
               2017, 2018, 2019, 2020, 2021)

the_sites <- rep('01-11-010', length(the_years))

Asheville.dfs <- map2(the_sites, the_years, scrape.it)
                       
Asheville.combined <- bind_rows(Asheville.dfs)  

ggplot(Asheville.combined, aes(x = Date, y = Max.Day.Use)) +
  geom_line() +
  geom_smooth(method = 'loess') +
  labs(x = "Date", y = "Average Daily Withdrawals", 
       title = 
         "Average Daily Withdrawals for 
       Asheville 2010 - 2021")

```

> Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?

> Asheville's water usage is increasing over time.
