---
title: "Assignment 5: Data Visualization"
author: "Allison Barbaro"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1.  Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Upload the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project, echo=TRUE}

#1 
library(tidyverse)
library(lubridate)
library(here)
library(cowplot)
library(ggplot2)

getwd()

Peter.Paul.nutrients <- 
  read.csv("Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
Niwo.Litter <- 
  read.csv("Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv")

#2 
Peter.Paul.nutrients$sampledate <-ymd(Peter.Paul.nutrients$sampledate)
Niwo.Litter$collectDate <-ymd(Niwo.Litter$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme, echo=TRUE}
#3
my_theme.1 <-
  theme(
    line = element_line(
      color='pink',
      linewidth =2
    ),
    legend.background = element_rect(
      color='grey',
    ),
    legend.title = element_text(
      color='blue'
    )
    
  )

theme(my_theme.1)

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4, echo=TRUE}
#4 
Phos.plot.1 <- Peter.Paul.nutrients %>%
  ggplot(aes(
    x=tp_ug,
    y=po4,
    color=lakename
  )) +
  geom_point() +
   labs(title = "Total Phosphorus by Phosphate",
       x ="Phosphorous", 
       y = "Phosphate") +
  geom_smooth(method=lm, color='black') +
  xlim(0, 70) +
  ylim(0, 50) + 
  scale_color_discrete(name = "Lake Name")
  
print(Phos.plot.1)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tip: R has a build in variable called `month.abb` that returns a list of months;see <https://r-lang.com/month-abb-in-r-with-example>

```{r Create boxplots, echo=TRUE}

#5 

Month.plot <- ggplot(Peter.Paul.nutrients) +
  geom_boxplot(aes(x= factor(month, levels=1:12, labels=month.abb), y = temperature_C, color = lakename)) +
  labs(x = "Month", y = "Temperature in Celsius", title = "Temperature of Peter and Paul Lake by Month") +
  scale_color_discrete(name = "Lake Name") + scale_x_discrete(drop=FALSE)
print(Month.plot)

M.P <- Month.plot + theme(legend.position = "none")

Month.plot.1 <- ggplot(Peter.Paul.nutrients) +
  geom_boxplot(aes(x= factor(month, levels=1:12, labels=month.abb), y = tp_ug, color = lakename)) +
  labs(x = "Month", y = "Total Phosphorus", title = "Phosphorous Levels of Peter and Paul Lake by Month") +
  scale_color_discrete(name = "Lake Name") + scale_x_discrete(drop=FALSE)
print(Month.plot.1)

M.P.1 <- Month.plot.1 + theme(legend.position = "none")

Month.plot.2 <- ggplot(Peter.Paul.nutrients) +
  geom_boxplot(aes(x= factor(month, levels=1:12, labels=month.abb), y = tn_ug, color = lakename)) + labs(x = "Month", y = "Total Nitrogen", title = "Nitrogen Levels of Peter and Paul Lake by Month") +
  scale_color_discrete(name = "Lake Name") + scale_x_discrete(drop=FALSE)
print(Month.plot.2)

M.P.2 <- Month.plot.2 + theme(legend.position = "none")

legend <- get_legend(Month.plot.1)

cowplot.main <- plot_grid(M.P, M.P.1, M.P.2, legend, ncol = 1)
print(cowplot.main)
  
```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: The variables are most present during the summer months, and largely absent during winter. Temperature is the variable that is most similar between the two lakes, while P and N are both more present in Peter Lake than Paul Lake.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter, echo=TRUE}
#6
Needles.Plot <- Niwo.Litter %>%
  filter(functionalGroup == "Needles") %>%
  ggplot(aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  geom_point() +
  labs(x = "Collection Date", y = "Dry Mass", title = "Needle Litter Dry Mass by Date") + scale_color_discrete(name = "NLCD Class")
  
print(Needles.Plot)

#7
Needles.Plot.facets <- Niwo.Litter %>%
  filter(functionalGroup == "Needles") %>%
  ggplot(aes(x = collectDate, y = dryMass)) +
  geom_point() +
  labs(x = "Collection Date", y = "Dry Mass", title = "Needle Litter Dry Mass by Date") +   scale_color_discrete(name = "NLCD Class") +
  facet_wrap(vars(nlcdClass))

print(Needles.Plot.facets)
  
```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: Plot 7 is more effective because the different NLCD classes are more clearly separated. Plot 6 there is too much overlap of data.